Community-Centric Innovation

### 2014-03-01 paper.tex
last day

\begin{abstract}
%\boldmath
The Internet is profoundly transforming the economic and social structures.
The ability to instantaneously share data across the globe enables new ways of collaboration and production models.
The individuals can join loosely organized working groups and communities and contribute to the evolution and progress of the humankind.
By choosing which communities they join and which communities they contribute to, the users can also be participants and producers.
Furthermore, this model enables a new way of democracy in which the people vote not only with their ballots, but also with their daily jobs and contributions.
In this work we describe examples of bottom-up initiatives in which the users play a central role in the development of the project.
These are initiatives by and for the people.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

@@

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{T}{he} growth of the Internet has dramatically accelerated the path of technological evolution.
The amount of information and knowledge available on the Net is increasing as well as the number of people that can access that knowledge.
Still, it is important to realize that the Internet is not simply an incremental evolution that brings faster progress.
The Internet is a disruptive innovation that is completely transforming the way we organize ourselves and relate to each other.

The fact that the information can be shared with little restrictions in terms of space, time and cost, makes it possible for people to organize in horizontal structures.
In this horizontal models, also known as peer-to-peer, all the participants can access to the information required to contribute to the decision-making process.
The heterogeneous background of the participants makes it possible to consider a wider space of solutions and the extensive scrutiny throughout the discussion and decision-making stage reduces the chances of mistakes.

@@

Another aspect that can draw the attention to community networks is privacy.
As commercial networks are routinely involved in mass surveillance programs, the citizens see in community networks a way to opt out from privacy-invading campaigns.
Finally, community networks are also a means of getting around censorship.
Totalitarian regimes build censorship into the networks to control which information is distributed among the population.
Filtering the information and ordering Internet blackouts have been used as ways for hindering social movements and placate population unrest.

@@

Mesh technologies such as the \emph{quick mesh project} (qmp) are evolving to make it easier for regular citizens to set up a network in a matter of minutes.
Friendly web interfaces for configuration abstract the underlying technological complexity.
In community networks local people become familiar with the procedures required to set up, maintain and troubleshoot the network.
In case of emergency, the presence of local experts facilitates quick deployments and reparations.

The decentralized nature of community networks greatly helps in fostering organic growth and resisting node failures.
The absence of central points of failure guarantees that two neighbours will be able to connect to each other as long as they keep their nodes running, regardless of what happens to the rest of the network.
The dynamic routing protocols, such as BMX6, allow the network to quickly re-configure itself and route packets around failures to keep connectivity.
To take a community network down it is necessary to shut down all the nodes.

Community networks are often less stable and more dynamic that their commercial counterparts.
The participants of community networks are eager to experiment and evolve all aspects of the network even if this results in occasional connectivity cuts.
As an example, while commercial networks seem to have a hard time transitioning to IPv6, qmp networks are native IPv6 networks.
Protocols, firmware and topology change on a daily basis to fix bugs, improve the performance or adapt to new situations.

This continuous change that forces to keep the equipment firmware updated represents  an additional layer of protection against spying agencies interested in compromising the network.
It is much more difficult to attack up-to-date system that those that run old unpatched firmwares.

@@

The participation in a community network trains the citizens in self-management techniques that inherit many of the properties of decentralized networks.
The lack of a central authority obliges the participants to take responsibility, exercise their freedom respecting the others and making concessions in order to achieve common goals.
%The organizational structures that we find in community networks mimic those of free (libre) software communities and other Internet-based coordinated projects such as Wikipedia.

@@

There are also examples of success of community-centric models in hardware initiatives.
Both Arduino and Raspberry-Pi are open hardware projects that are aligned with community-centric innovation.
They both share all the information regarding the hardware with the community of users that, in turn, use this information to contribute to the project.

>>>>

\begin{abstract}
%\boldmath
The Internet is profoundly transforming the economic and social structures of the 21st century.
The ability to instantaneously share data across the globe enables new ways of collaboration and production models.
The individuals can join loosely organized working groups and communities and contribute to the evolution and progress of the humankind.
By choosing which communities they join and which communities they contribute to, the users can also be participants and producers.
Furthermore, this model enables a new way of democracy in which the people vote not only with their ballots, but also with their daily jobs and contributions.
In this work we describe examples of bottom-up initiatives in which the users play a central role in the development of the project.
These are initiatives by and for the people.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

@@

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{T}{he} growth of the Internet has dramatically accelerated the path of technological evolution.
The amount of information and knowledge available on the Net is increasing.
Importantly,  the number of people that can access that knowledge is increasing as well.
The Internet is more than an incremental evolution that brings faster progress.
It is a disruptive innovation that is completely transforming the way we organize ourselves and relate to each other.

The fact that the information can be shared with little restrictions in terms of space, time and cost, makes it possible for people to organize in horizontal structures.
In this horizontal models, also known as peer-to-peer models, all the participants can access to the information required to contribute to the decision-making process.
By removing the barriers for participation, the working force of peer-to-peer projects is reinforced.
The benefit is not only the number of participants, but also the variety.
The heterogeneous background of the participants makes it possible to consider a wider space of solutions and the extensive scrutiny throughout the discussion and decision-making stage reduces the chances of mistakes.

@@

Another aspect that can draw the attention to community networks is privacy.
As commercial networks are involved in mass surveillance programs, the citizens see in community networks a way to opt out from privacy-invading campaigns.
Finally, community networks are also a means of getting around censorship.
Totalitarian regimes build censorship into the networks to control which information is distributed among the population.
Filtering the information and ordering Internet blackouts have been used as ways for hindering social movements and placate population unrest.

@@

Mesh technologies such as the \emph{quick mesh project} (qmp) are evolving to make it easier for regular citizens to set up a network in a matter of minutes.
Friendly web interfaces for configuration abstract the underlying technological complexity.
In community networks, local people become familiar with the procedures required to set up, maintain and troubleshoot the network.
In case of emergency, the presence of local experts facilitates quick deployments and reparations.

The decentralized nature of community networks greatly helps in fostering organic growth and also in  resisting node failures.
The absence of central points of failure guarantees that two neighbours will be able to connect to each other as long as they keep their nodes running, regardless of what happens to the rest of the network.
Mesh routing protocols, such as BMX6, allow the network to quickly re-configure itself and route packets around failures to keep connectivity.
These protocols are particularly tailored to work in challenging multi-hop wireless conditions.
To take down a community network it is necessary to shut down all the nodes.

Community networks are often less stable and more dynamic than their commercial counterparts.
The participants of community networks are eager to experiment and evolve all aspects of the network even if this results in an increased risk of occasional connectivity cuts.
As an example, while commercial networks run mostly on IPv4, qmp networks are native IPv6 networks.
Protocols, firmware and topology change on a daily basis to fix bugs, improve the performance or adapt to new situations.

This continuous change that forces to keep the equipment firmware updated represents  an additional layer of protection against spying agencies interested in compromising the network.
It is much more difficult to attack up-to-date system than it is to attack systems running old unpatched firmwares.

@@

The participation in a community network trains the citizens in self-management techniques that inherit many of the properties of decentralized networks.
The lack of a central authority obliges the participants to take responsibility, exercise their freedom and respect the others.
Making concessions is necessary in order to achieve common goals.
%The organizational structures that we find in community networks mimic those of free (libre) software communities and other Internet-based coordinated projects such as Wikipedia.

@@

There are also examples of success of community-centric models in hardware initiatives.
Both Arduino and BeagleBoard are open hardware projects that are aligned with community-centric innovation.
They both share all the information regarding the hardware with the community of users that, in turn, use this information to contribute to the project.
